Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool

### 2024-10-05 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
update rev 2

The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
This information is crucial for an accurate PT trip and route estimation.
A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing. <!--  with a 25 m spatial resolution, -->
The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.

@@

The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
This map shows that some of the municipalities are not served by PT modes that can easily accommodate bicycles, which may limit the potential for intermodality in these areas.

This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
This restrictions are based on the assumption that 5 km is a plausible maximum distance that people are willing to travel by bike as part of a longer journey.
This applies especially for cycling beginners shifting from cars, who are less likely to be confident and experienced cyclists than the existing cyclists.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles, at an aggregate level of the LMA. 
For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling, even given the travel restrictions considered (up to 5 km on bike, up to 2 hours, no possible transfers between PT).
This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).

@@

knitr::kable(summary21 |> select(-Baseline),
      digits = 0,
      format.args = list(big.mark = " "),
      # col.names = c("Target", "Routing", "Total",
      #               "Bus", "Ferry", "Train", "Tram"),
      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(1:7), width = "4.5em")

@@

Another relevant aspect is that the higher potential (Figure \ref{fig:mapzones}.a) is in the municipalities with the highest number of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and Almada, which are also the ones with the highest number of PT interfaces (Figure \ref{fig:map1}).
On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setubal. 
This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setubal).

>>>>

Trips were routed for all the OD jittered pairs, resulting in the least-cost paths (shortest travel time) for each pair.

Figure \ref{fig:trip_schematic} illustrates a schematic comparison between *safe* and *direct* routing profiles, showing both bicycle-only and Bike + PT options.
In the *direct* profile, bicycle-only trips can be more attractive due to shorter travel times, especially for shorter distances, although, for longer trips, combining cycling with PT can be faster.
In the *safe* routing profile, where longer, less stressful routes (LTS 3 or lower), are used to avoid high-stress roads (LTS 4), cycling in combination with PT can be more efficient.

```{r trip_schematic, fig.align='center', fig.cap="Schematic trip for safe and direct routing profiles, with bicycle-only and bicycle with PT options. The black lines represent the Biclcye legs, and the dashed orange lines represent the PT leg.", fig.height=4, out.width="70%"}
# png::writePNG(pdftools::pdf_render_page("img/schematic_trips.pdf", dpi = 300), "img/schematic_trips_cairo.png")
knitr::include_graphics("img/schematic_trips_cairo.png", error = FALSE)
```

The `r5r` model uses the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
This information is crucial for an accurate PT trip and route estimation.
A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing. <!--  with a 25 m spatial resolution, -->
The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.

@@

The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.

We have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
This map shows that some of the municipalities are not served by PT modes that can easily accommodate bicycles, which may limit the potential for intermodality in these areas.

Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
Additionally, we restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).

<!-- This resulted in that only 53.8% of the total trips between the jittered origins and destinations are possible to be made by bicycle-only or bicycle + PT (for the considered PT modes) for the _safe_ routing profile, and 69.2% for the _direct_ routing profile. -->

<!-- After the estimation of routes, we considered only the Origin-Destination pairs for which the estimated routes were the ones that were more attractive (least-cost path / fastest) for the bicycle and PT combination, than for bicycle only. -->

This conservative approach, regarding the assumed multi-modal restrictions, was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
This restrictions are based on the assumption that 5 km is a plausible maximum distance that people are willing to travel by bike as part of a longer journey.
This applies especially for cycling beginners shifting from cars, who are less likely to be confident and experienced cyclists than the existing cyclists.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles, at an aggregate level of the LMA. 
It should be noted that for the OD pairs with already 4% or 10% bike trips in 2018, those are not accounted as new potential bike trips, resulting in potential Cycling + PT numbers slightly below the 4% and 10% of the total trips.

For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.

@@

Even given the travel restrictions considered (up to 5 km on bike, up to 2 hours, no possible transfers between PT), this resulted in 538.514 trips with potential to be made by bicycle in combination with PT (`r round(538514/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips) for the *safe* routing profile, and 500.880 trips for the *direct* routing profile (`r round(500880/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips).
This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).

@@

knitr::kable(summary21 |> select(-Baseline),
      digits = 0,
      format.args = list(big.mark = " "),
      # col.names = c("Target", "Routing", "Total",
      #               "Bus", "Ferry", "Train", "Tram"),
      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disaggregated by PT mode. Values in 'trips/day'.",
      format="latex",
      booktabs = TRUE) |> 
  kableExtra::column_spec(column = c(1:7), width = "4.5em")

@@

The higher number of Bike + PT trips in the _safe_ routing profile can be explained by the reduced competitiveness of cycling alone in this profile, as it restricts cyclists to lower-traffic sections, with less directed options, and a subset of the total road network.
In contrast, the _direct_ routing profile, which allows for direct cycling routes along the entire network, often results in faster single-mode bicycle trips - more competitive than using the Bike in combination with PT.
As a result, more trips are made using Bike + PT under the _safe_ routing profile, while _direct_ routes favour cycling-only journeys.
This justifies the aggregate difference in the number of potential trips between the two profiles.

For trams, which primarily operate along busy main roads that are generally hostile for cycling (classified as LTS 4), the Bike + Tram combination is often less competitive compared to direct bike-only routes, especially for shorter trips.
This is particularly evident in the case of Almada, the only municipality in this study with a tram system that can be integrated into Bike + PT trips (see Figure \ref{fig:map1}).
For _direct_ routing, bike-only options are usually faster for shorter local trips within the municipality.
However, under _safe_ routing assumptions (LTS 3), where cyclists are limited to less direct, low-stress routes, the Bike + Tram option becomes more attractive, as the bike-only option would be longer, albeit safer, avoiding the main roads.
This explains the higher competitiveness of Tram + Bike in the safe profile in this municipality.

Another relevant aspect is that the higher potential (Figure \ref{fig:mapzones}.a) is in the municipalities with the highest number of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and Almada, which are also the ones with the highest number of PT interfaces (Figure \ref{fig:map1}).
On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setúbal. 
This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setúbal).

<!-- A particular difference is noted regarding the attraction of trams to be used in combination with cycling, for _safe_ and _direct_ routes. -->
<!-- In this case, Tram is available for intermodality only at one municipality (Almada) (see \ref{fig:map1}), and most of its stations are located on major roads. -->
<!-- Their accessibility involves travelling along these major roads (with LTS 4), which limits the potential of valid routing options when considering _safe_ routes.  -->
<!-- In this case, Train offers the highest potential for _safe_ routes, and Tram for _direct_ routes. -->

### 2024-10-05 paper/PaperCEUS/PaperCEUS_rev2_word.tex
update rev 2

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data aggregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model,
from the European Space Agency's COPERNICUS mission, was used to include
street gradient information, as a weight in cycling routing. The cycling
potential trips for the two national strategic targets (4\% and 10\%)
were estimated from the 2017 cycling and car trips (both as a driver and
as a passenger), the baseline scenario.

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs. We restricted our analysis to the
first and last legs with a combined length of up to 5 km (for example: 1
km from origin to interface \emph{A} plus 4 km from interface \emph{B}
to destination). Furthermore, we have imposed restrictions on PT usage,
limiting it to trips without PT transfers, and within a duration of up
to 2 hours (120 minutes). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}). This map shows that some of the municipalities are not
served by PT modes that can easily accommodate bicycles, which may limit
the potential for intermodality in these areas.

This conservative approach was adopted to capture the fact that cycling
stages as part of a multi-modal trip are likely to be shorter than
cycling-only trips \citep{vanmil_insights_2021}. Additionally,
\citet{Leferink2017} found that the bicycle is most attractive option as
a first and last mile mode when the distance ranges from 3 to 5 km, as
observed across various comparative studies. This restrictions are based
on the assumption that 5 km is a plausible maximum distance that people
are willing to travel by bike as part of a longer journey. This applies
especially for cycling beginners shifting from cars, who are less likely
to be confident and experienced cyclists than the existing cyclists.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles, at an aggregate level of the LMA. For the
cycling legs of the journey (first and last legs), the environmental
avoided emissions and monetized socio-environment (SE) benefits are
presented in Table \ref{tab:summary1b}, resulting from replacing car
trips with cycling.

@@

For both \emph{direct} and \emph{safe} route profiles, 10\% of the daily
trips have the potential to be made by a combination of PT and cycling,
even given the travel restrictions considered (up to 5 km on bike, up to
2 hours, no possible transfers between PT). This unveils the potential
of cycling as a complementary mode of PT, with the potential to uptake
the number of PT trips within the LMA area by as much as 6.3\% (in
addition to the 825 thousand PT trips reported in the mobility survey).

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

Another relevant aspect is that the higher potential (Figure
\ref{fig:mapzones}.a) is in the municipalities with the highest number
of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and
Almada, which are also the ones with the highest number of PT interfaces
(Figure \ref{fig:map1}). On another hand, the municipalities with little
PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
lower potential for car to Bike-PT substitution (Figure
\ref{fig:mapzones}.a), even if they have a high number of trips, such as
Setubal. This findings are aligned with their local travel patterns,
where the majority of trips are intra-municipal (67\% for both
municipalities), and not to the city of Lisbon, despite the high
car-dependency (79\% in Mafra, 67\% in Setubal).

>>>>

Trips were routed for all the OD jittered pairs, resulting in the
least-cost paths (shortest travel time) for each pair.

Figure \ref{fig:trip_schematic} illustrates a schematic comparison
between \emph{safe} and \emph{direct} routing profiles, showing both
bicycle-only and Bike + PT options. In the \emph{direct} profile,
bicycle-only trips can be more attractive due to shorter travel times,
especially for shorter distances, although, for longer trips, combining
cycling with PT can be faster. In the \emph{safe} routing profile, where
longer, less stressful routes (LTS 3 or lower), are used to avoid
high-stress roads (LTS 4), cycling in combination with PT can be more
efficient.

{\centering \includegraphics[width=0.7\linewidth,]{img/schematic_trips_cairo}

\caption{Schematic trip for safe and direct routing profiles, with bicycle-only and bicycle with PT options. The black lines represent the Biclcye legs, and the dashed orange lines represent the PT leg.}\label{fig:trip_schematic}
\end{figure}

The \texttt{r5r} model uses the OpenStreetMap road network and the GTFS
metropolitan data aggregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model,
from the European Space Agency's COPERNICUS mission, was used to include
street gradient information, as a weight in cycling routing. The cycling
potential trips for the two national strategic targets (4\% and 10\%)
were estimated from the 2017 cycling and car trips (both as a driver and
as a passenger), the baseline scenario.

@@

The intermodality scenario considers trips made by PT in which cycling
is used for the first and last legs.

We have only included PT modes that can easily accommodate bicycles,
such as trains, ferries, trams, and inter-municipal bus lines equipped
with bike racks (Figure \ref{fig:map1}). This map shows that some of the
municipalities are not served by PT modes that can easily accommodate
bicycles, which may limit the potential for intermodality in these
areas.

Furthermore, we have imposed restrictions on PT usage, limiting it to
trips without PT transfers, and within a duration of up to 2 hours (120
minutes). Additionally, we restricted our analysis to the first and last
legs with a combined length of up to 5 km (for example: 1 km from origin
to interface \emph{A} plus 4 km from interface \emph{B} to destination).

This conservative approach, regarding the assumed multi-modal
restrictions, was adopted to capture the fact that cycling stages as
part of a multi-modal trip are likely to be shorter than cycling-only
trips \citep{vanmil_insights_2021}. Additionally, \citet{Leferink2017}
found that the bicycle is most attractive option as a first and last
mile mode when the distance ranges from 3 to 5 km, as observed across
various comparative studies. This restrictions are based on the
assumption that 5 km is a plausible maximum distance that people are
willing to travel by bike as part of a longer journey. This applies
especially for cycling beginners shifting from cars, who are less likely
to be confident and experienced cyclists than the existing cyclists.

@@

Table \ref{tab:summary1} presents the LMA total daily trips that can be
made with cycling + TP combination (with the aforementioned
restrictions), the trips in the baseline scenario and corresponding new
daily trips to achieve the national strategy targets (4\% and 10\%), for
different route profiles, at an aggregate level of the LMA. It should be
noted that for the OD pairs with already 4\% or 10\% bike trips in 2018,
those are not accounted as new potential bike trips, resulting in
potential Cycling + PT numbers slightly below the 4\% and 10\% of the
total trips.

For the cycling legs of the journey (first and last legs), the
environmental avoided emissions and monetized socio-environment (SE)
benefits are presented in Table \ref{tab:summary1b}, resulting from
replacing car trips with cycling.

@@

Even given the travel restrictions considered (up to 5 km on bike, up to
2 hours, no possible transfers between PT), this resulted in 538.514
trips with potential to be made by bicycle in combination with PT
(10.1\% of all daily trips) for the \emph{safe} routing profile, and
500.880 trips for the \emph{direct} routing profile (9.4\% of all daily
trips). This unveils the potential of cycling as a complementary mode of
PT, with the potential to uptake the number of PT trips within the LMA
area by as much as 6.3\% (in addition to the 825 thousand PT trips
reported in the mobility survey).

@@

\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disaggregated by PT mode. Values in 'trips/day'.}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
\toprule
Target & Routing & Potential & Bus & Ferry & Train & Tram\\
\midrule
4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
\bottomrule
\end{tabular}
\end{table}

The higher number of Bike + PT trips in the \emph{safe} routing profile
can be explained by the reduced competitiveness of cycling alone in this
profile, as it restricts cyclists to lower-traffic sections, with less
directed options, and a subset of the total road network. In contrast,
the \emph{direct} routing profile, which allows for direct cycling
routes along the entire network, often results in faster single-mode
bicycle trips - more competitive than using the Bike in combination with
PT. As a result, more trips are made using Bike + PT under the
\emph{safe} routing profile, while \emph{direct} routes favour
cycling-only journeys. This justifies the aggregate difference in the
number of potential trips between the two profiles.

For trams, which primarily operate along busy main roads that are
generally hostile for cycling (classified as LTS 4), the Bike + Tram
combination is often less competitive compared to direct bike-only
routes, especially for shorter trips. This is particularly evident in
the case of Almada, the only municipality in this study with a tram
system that can be integrated into Bike + PT trips (see Figure
\ref{fig:map1}). For \emph{direct} routing, bike-only options are
usually faster for shorter local trips within the municipality. However,
under \emph{safe} routing assumptions (LTS 3), where cyclists are
limited to less direct, low-stress routes, the Bike + Tram option
becomes more attractive, as the bike-only option would be longer, albeit
safer, avoiding the main roads. This explains the higher competitiveness
of Tram + Bike in the safe profile in this municipality.

Another relevant aspect is that the higher potential (Figure
\ref{fig:mapzones}.a) is in the municipalities with the highest number
of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and
Almada, which are also the ones with the highest number of PT interfaces
(Figure \ref{fig:map1}). On another hand, the municipalities with little
PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a
lower potential for car to Bike-PT substitution (Figure
\ref{fig:mapzones}.a), even if they have a high number of trips, such as
Setúbal. This findings are aligned with their local travel patterns,
where the majority of trips are intra-municipal (67\% for both
municipalities), and not to the city of Lisbon, despite the high
car-dependency (79\% in Mafra, 67\% in Setúbal).
